# Correlation coefficients and working with normal distributions {#sec-ISACh3}

## Learning goals

### Conceptual

1. Calculate and interpret pearson's $r$ by hand
2. Z-transform datapoints and interpret the result

### SPSS

1. Compute pearson's $r$

## Pearson's $r$ correlation coefficient

Pearson's $r$ is given by:

<br>  

$$
r_{xy} = \frac{\sum_{i=1}^{n}{(x_i - \bar{x})(y_i - \bar{y})}}{{\sqrt{\sum_{i=1}^{n}{(x_i - \bar{x})^2}}}{\sqrt{\sum_{i=1}^{n}{(y_i - \bar{y})^2}}}}
$$ 
<br>  

Where:

* $x_i$ = i'th score on variable $x$  
* $y_i$ = i'th score on variable $y$  
* $\bar{x}$ = mean of variable $x$  
* $\bar{y}$ = mean of variable $y$  
* $\sum_{i=1}^{n}$ = sum scores from index (element) 1 to $n$   
* $n$ = sample size

Let's now imagine that we develop the following Research Question:_Does the average daily time (in minutes) spent using SPSS correlate with the average number of obscenities used per day?_ We collect some data, which looks like this:

<br>  

```{r}
#| label: tbl-corrdat
#| tbl-cap: "Average SPSS minutes per week vs. average obscenities uttered"
#| tbl-cap-location: bottom 

`%>%`<-magrittr::`%>%`  

data<-tibble::tibble("Average SPSS minutes"=c(240,310,230,220),"Average # obscenities"=c(3,6,4,3)) %>%
  knitr::kable(align="c")
data
```


<br>  

With the following steps, we can unpack the formula and derive the correlation coefficient:

<br>  

```{r}
#| label: tbl-corrdat_unpack
#| tbl-cap: "Correlation steps"
#| tbl-cap-location: bottom 

data<-tibble::tibble("minutes_x"=c(240,310,230,220),"words_y"=c(3,6,4,3))
data$x_minus_xbar<-data$minutes_x-mean(data$minutes_x)
data$y_minus_ybar<-data$words_y-mean(data$words_y)
data$xdev_by_ydev<-data$x_minus_xbar*data$y_minus_ybar
data$xdev_square<-data$x_minus_xbar^2
data$ydev_square<-data$y_minus_ybar^2

# data %>%
#   knitr::kable(align="c",col.names=c("$x_{minutes}$","$y_{words}$","$x-\\bar{x}$","$y-\\bar{y}$","$(x-\\bar{x}).(y-\\bar{y})$","$(x-\\bar{x})^2$","$(y-\\bar{y})^2$"))

data %>%
  kableExtra::kbl(align="c",col.names=c("$$x_{minutes}$$","$$y_{words}$$","$$x-\\bar{x}$$","$$y-\\bar{y}$$","$$(x-\\bar{x}).(y-\\bar{y})$$","$$(x-\\bar{x})^2$$","$$(y-\\bar{y})^2$$")) %>% 
  kableExtra::kable_classic() %>% 
  kableExtra::column_spec(column=1,background="white") %>% 
  kableExtra::column_spec(column=2,background="lightblue") %>%
  kableExtra::column_spec(column=3,background="navajowhite") %>%
  kableExtra::column_spec(column=4,background="lightsalmon") %>%
  kableExtra::column_spec(column=5,background="lightgreen") %>% 
  kableExtra::column_spec(column=6,background="lightpink") %>% 
  kableExtra::column_spec(column=7,background="grey")
```

<br>  

* $\bar{x} = 250$   
* $\bar{y}= 4$     
* $\sum_{i=1}^{n}{(x_i - \bar{x})(y_i - \bar{y})}=160$   
* $\sum_{i=1}^{n}{(x_i - \bar{x})^2} = 5000$  
* $\sum_{i=1}^{n}{(y_i - \bar{y})^2} = 6$

<br>  

$$
r_{xy} = \frac{160}{\sqrt{5000}*\sqrt{6}} = \frac{160}{173.21} = .92
$$
<br>  

We will use the following rules-of-thumb for interpreting **absolute** values of $r$:

* $r \le .3$ = small  
* $.3 < r \le .6$ = moderate    
* $.6 < r \le .8$ = strong  
* $.8 < r \le 1$ =  very strong 

## Z-transformations

Imagine we know that income per week (in euro) is normally distributed in the population of Dutch university students who graduated in the last year, such that the mean ($\mu$) income is  1,500 EUR, with an SD ($\sigma$) of 100. We might wonder: _what proportion of the population of new Dutch graduates would have a weekly income of greater than 1,700 EUR?_

```{r}
#| label: fig-normdist
#| fig-cap: "Fictional distribution of graduate income scores. Green shading denotes incomes of greater than 1,700 EUR."
#| echo: FALSE 

x<-seq(1100,1900,10)
y<-dnorm(x,1500,100)
xx<-tibble::tibble("Income"=x,"pdf"=y)
mulab<- paste("mu[income] == ", 1500)
siglab<- paste("sigma[income] == ", 100)
p <- ggplot2::ggplot(xx, ggplot2::aes(x=Income,y=pdf)) +
  ggplot2::geom_line()+
  ggplot2::geom_segment(x=1500,y=0,xend=1500,yend=max(xx$pdf),linetype="dashed",color="red")+
  ggplot2::geom_segment(x=1400,y=0,xend=1400,yend=as.numeric(xx[xx$Income==1400,2]),linetype="dashed",color="blue")+
  ggplot2::geom_segment(x=1600,y=0,xend=1600,yend=as.numeric(xx[xx$Income==1600,2]),linetype="dashed",color="blue")+
  ggplot2::geom_segment(x=1300,y=0,xend=1300,yend=as.numeric(xx[xx$Income==1300,2]),linetype="dashed",color="blue")+
  ggplot2::geom_segment(x=1700,y=0,xend=1700,yend=as.numeric(xx[xx$Income==1700,2]),linetype="dashed",color="blue")+
  ggplot2::annotate("text", x = 1150, y = 0.0041, colour="red", label = mulab,parse=TRUE)+
  ggplot2::annotate("text", x = 1150, y = 0.0039, colour="blue", label = siglab,parse=TRUE)+
  ggplot2::geom_ribbon(data=subset(xx,x>=1700),ggplot2::aes(x=Income,ymax=pdf),ymin=0,fill="green")+
  ggplot2::scale_x_continuous(breaks = seq(1100,1900,100),labels = seq(1100,1900,100))+
  ggplot2::theme(axis.text.y = ggplot2::element_blank())+
  ggplot2::ylab("Density")
p
```

We can see the area of interest in the distribution above, in @fig-normdist. To calculate the proportion, we first need to convert the target value into a **z-score**, which transforms the value into standard deviation units. In this case:

<br>  

$$
z = \frac{x-\mu}{\sigma} = \frac{1700-1500}{100} = 2
$$
```{r}
target_income<-1700
mean_income<-1500
sd_income<-100

prob<-pnorm(target_income ,mean = mean_income, sd = sd_income, lower.tail = FALSE)

# We set 'lower.tail = FALSE' because we want to return probabilities associated with a score *greater* than 1,700, so the upper/right-tail values.

```

<br>  

To work out the probability associated with a z-score of 2, we can consult the normal distribution tables in the Argyrous textbook (pp. 541 - 542), looking for the area under the curve <b>beyond</b> one point. This allows us to answer our question: _there is a `{r} round(prob*100,2)`% chance of randomly selecting an individual with an income above 1,700 EUR per week from the population of new Dutch graduates_

## Scatterplots and $r$ with the WDI data (SPSS)

Let's again load in the World Data Indicators (WDI) dataset (`World_Data.sav`), and revisit our scatterplot from the previous class (see @fig-wdscatter_1).

```{r}
#| eval: FALSE

wd<-haven::read_sav("World_Data.sav")
wd %>% 
  ggplot2::ggplot(ggplot2::aes(x = wdi_water, y = wdi_lifexp)) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(method = "lm", se = FALSE)
```

```{r}
#| echo: FALSE
#| label: fig-wdscatter_ISA3
#| fig-cap: "Access to water vs. Life exp."
#| warning: FALSE
#| message: FALSE

wd<-haven::read_sav("C:/Users/David/OneDrive - UvA/2025_2026/ISA_24_25/Datasets/World_Data.sav")

wd %>% 
  ggplot2::ggplot(ggplot2::aes(x = wdi_water, y = wdi_lifexp)) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(method = "lm", se = FALSE)
```

We see a noisy, yet positive relationship between a country's percentage access to an improved water source and its life expectancy at birth. The points do not cling tightly around the linear fit line, but are scattered around it.

```{r}

wdcor <- cor(wd$wdi_water, wd$wdi_lifexp, use = "complete.obs")

```

To give us some more formal information about the relationship strength, we can calculate pearson's $r$ in SPSS which reveals that the correlation between the variables is `{r} round(wdcor,2)`. Following our rules-of-thumb above, we could diagnose this as a **moderate, positive** relationship between the two variables.


## SPSS explainer video

[WDI scatterplot, pearson]
